### 2025-04-09 correlation.qmd
20250409 - clean up and cross references

Correlation is an index of the association between variables.
Covariance is the association between variables and in an unstandardized metric that differs for variables with different scales.
By contrast, correlation is in a standardized metric that does not differ for variables with different scales.
When examining the association between variables that are [interval](#sec-interval) or [ratio](#sec-ratio) levels of measurement, Pearson correlation is used.
When examining the association between variables that are [ordinal](#sec-ordinal) in level of measurement, Spearman correlation is used.
Pearson correlation is an index of the *linear* association between variables.
If a nonlinear association is present, other indices like xi [$\xi$; @Chatterjee2021] and distance correlation coefficients are better suited to detect the association.

@@

Keep in mind that the ([Pearson](#sec-correlationExamplesPearson)) correlation examines the strength of the *linear* association between two variables.\index{correlation}
If the association between two variables is nonlinear, the ([Pearson](#sec-correlationExamplesPearson)) correlation provides the strength of the linear trend and may not provide a meaningful index of the strength of the association between the variables.\index{correlation}
For instance, Anscombe's quartet includes four sets of data that have nearly identical basic descriptive statistics (see Tables [-@tbl-anscombe] and [-@tbl-anscombeStats]), including the same bivariate correlation, yet have very different distributions and whose association takes very different forms (see Figure \@ref(fig:anscombeQuartet)).\index{correlation}

| x1 | y1    | x2 | y2   | x3 | y3    | x4 | y4   |
|----|-------|----|------|----|-------|----|------|
| 10 | 8.04  | 10 | 9.14 | 10 | 7.46  | 8  | 6.58 |
| 8  | 6.95  | 8  | 8.14 | 8  | 6.77  | 8  | 5.76 |
| 13 | 7.58  | 13 | 8.74 | 13 | 12.74 | 8  | 7.71 |
| 9  | 8.81  | 9  | 8.77 | 9  | 7.11  | 8  | 8.84 |
| 11 | 8.33  | 11 | 9.26 | 11 | 7.81  | 8  | 8.47 |
| 14 | 9.96  | 14 | 8.1  | 14 | 8.84  | 8  | 7.04 |
| 6  | 7.24  | 6  | 6.13 | 6  | 6.08  | 8  | 5.25 |
| 4  | 4.26  | 4  | 3.1  | 4  | 5.39  | 19 | 12.5 |
| 12 | 10.84 | 12 | 9.13 | 12 | 8.15  | 8  | 5.56 |
| 7  | 4.82  | 7  | 7.26 | 7  | 6.42  | 8  | 7.91 |
| 5  | 5.68  | 5  | 4.74 | 5  | 5.73  | 8  | 6.89 |

@@

| Property                     | Value        |
|------------------------------|--------------|
| Sample size                  | 11           |
| Mean of X                    | 9.0          |
| Mean of Y                    | ~7.5         |
| Variance of X                | 11.0         |
| Variance of Y                | ~4.1         |
| Equation of regression line  | Y = 3 + 0.5X |
| Standard error of slope      | 0.118        |
| One-sample t-statistic       | 4.24         |
| Sum of squares of X          | 110.0        |
| Regression sum of squares    | 27.50        |
| Residual sum of squares of Y | 13.75        |
| Correlation coefficient      | .816         |
| Coefficient of determination | .67          |

@@

@@

anscombe

@@

The covariance is an unstandardized index of the association between two variables.
However, it can be helpful to use a standardized index to examine the [effect size](#sec-practicalSignificance)—i.e., the strength of the association.
The [Pearson correlation coefficient](Pearson Correlation), $r$, is an example of a standardized index of the covariance between two variables.

@@

The [`petersenlab`](https://cran.r-project.org/web/packages/petersenlab/index.html) package [@R-petersenlab] contains the `cor.table()` function that generates a correlation matrix of variables, including the $r$ value, number of observations ($n$), asterisks denoting statistical significance, and $p$-value for each association.
The asterisks follow the following traditional convention for statistical significance: $^\dagger p < .1$; $^* p < .05$; $^{**} p < .01$; $^{***} p < .001$.
Here is a Pearson correlation matrix:

@@

Here is a Spearman correlation matrix:

@@

The [`petersenlab`](https://cran.r-project.org/web/packages/petersenlab/index.html) package [@R-petersenlab] contains the `partialcor.table()` function that generates a partial correlation matrix.
Here is a partial correlation matrix, controlling for the player's age:

@@

```{r}
#| label: fig-plotWithOutlier
#| fig-cap: "Scatterplot With Outlier."
#| fig-alt: "Scatterplot With Outlier."
#| code-fold: true

@@

Note how the association was not close to being statistically significant with the outlier, but with the outlier added, the association is statistically significant.
One way to combat this is to use methods for estimating the association between variables that are robust to (i.e., less impacted by) outliers.
In the next section, we describe robust correlation methods.

>>>>

Correlation is an index of the association between variables.
Covariance is the association between variables and in an unstandardized metric that differs for variables with different scales.
By contrast, correlation is in a standardized metric that does not differ for variables with different scales.
When examining the association between variables that are [interval](#sec-interval) or [ratio](#sec-ratio) levels of measurement, [Pearson correlation](#sec-correlationExamplesPearson) is used.
When examining the association between variables that are [ordinal](#sec-ordinal) in level of measurement, [Spearman correlation](#sec-correlationExamplesSpearman) is used.
[Pearson correlation](#sec-correlationExamplesPearson) is an index of the *linear* association between variables.
If a nonlinear association is present, other indices like xi [$\xi$; @Chatterjee2021] and distance correlation coefficients are better suited to detect the association.

@@

Keep in mind that the ([Pearson](#sec-correlationExamplesPearson)) correlation examines the strength of the *linear* association between two variables.\index{correlation}
If the association between two variables is nonlinear, the ([Pearson](#sec-correlationExamplesPearson)) correlation provides the strength of the linear trend and may not provide a meaningful index of the strength of the association between the variables.\index{correlation}
For instance, Anscombe's quartet includes four sets of data that have nearly identical basic descriptive statistics (see Tables [-@tbl-anscombe] and [-@tbl-anscombeStats]), including the same bivariate correlation, yet have very different distributions and whose association takes very different forms (see @fig-anscombeQuartet).\index{correlation}

| x1 | y1    | x2 | y2   | x3 | y3    | x4 | y4    |
|---:|------:|---:|-----:|---:|------:|---:|------:|
| 10 | 8.04  | 10 | 9.14 | 10 | 7.46  | 8  | 6.58  |
| 8  | 6.95  | 8  | 8.14 | 8  | 6.77  | 8  | 5.76  |
| 13 | 7.58  | 13 | 8.74 | 13 | 12.74 | 8  | 7.71  |
| 9  | 8.81  | 9  | 8.77 | 9  | 7.11  | 8  | 8.84  |
| 11 | 8.33  | 11 | 9.26 | 11 | 7.81  | 8  | 8.47  |
| 14 | 9.96  | 14 | 8.10 | 14 | 8.84  | 8  | 7.04  |
| 6  | 7.24  | 6  | 6.13 | 6  | 6.08  | 8  | 5.25  |
| 4  | 4.26  | 4  | 3.10 | 4  | 5.39  | 19 | 12.50 |
| 12 | 10.84 | 12 | 9.13 | 12 | 8.15  | 8  | 5.56  |
| 7  | 4.82  | 7  | 7.26 | 7  | 6.42  | 8  | 7.91  |
| 5  | 5.68  | 5  | 4.74 | 5  | 5.73  | 8  | 6.89  |

@@

| Property                     | Value        |
|:-----------------------------|:-------------|
| Sample size                  | 11           |
| Mean of X                    | 9.0          |
| Mean of Y                    | ~7.5         |
| Variance of X                | 11.0         |
| Variance of Y                | ~4.1         |
| Equation of regression line  | Y = 3 + 0.5X |
| Standard error of slope      | 0.118        |
| One-sample t-statistic       | 4.24         |
| Sum of squares of X          | 110.0        |
| Regression sum of squares    | 27.50        |
| Residual sum of squares of Y | 13.75        |
| Correlation coefficient      | .816         |
| Coefficient of determination | .67          |

@@

```{r}
#| output: false

anscombe
```

@@

@@

The covariance is an unstandardized index of the association between two variables.
However, it can be helpful to use a standardized index to examine the [effect size](#sec-practicalSignificance)—i.e., the strength of the association.
The [Pearson correlation coefficient](#sec-correlationExamplesPearson), $r$, is an example of a standardized index of the covariance between two variables.

@@

The [`petersenlab`](https://cran.r-project.org/web/packages/petersenlab/index.html) package [@R-petersenlab] contains the `cor.table()` function that generates a correlation matrix of variables, including the $r$ value, number of observations ($n$), asterisks denoting statistical significance, and $p$-value for each association.
The asterisks follow the following traditional convention for statistical significance: $^\dagger p < .1$; $^* p < .05$; $^{**} p < .01$; $^{***} p < .001$.
Here is a [Pearson correlation](#sec-correlationExamplesPearson) matrix:

@@

Here is a [Spearman correlation](#sec-correlationExamplesSpearman) matrix:

@@

The [`petersenlab`](https://cran.r-project.org/web/packages/petersenlab/index.html) package [@R-petersenlab] contains the `partialcor.table()` function that generates a [partial correlation](#sec-correlationExamplesPartial) matrix.
Here is a [partial correlation](#sec-correlationExamplesPartial) matrix, controlling for the player's age:

@@

```{r}
#| label: fig-plotWithOutlier
#| fig-cap: "Scatterplot With Outlier (in Red)."
#| fig-alt: "Scatterplot With Outlier (in Red)."
#| code-fold: true

@@

Note how the association was not close to being statistically significant without the outlier, but with the outlier added, the association is statistically significant.
One way to combat this is to use methods for estimating the association between variables that are robust to (i.e., less impacted by) outliers.
In the next section, we describe robust correlation methods.
